### 2017-03-25 thesis/results/results.tex
finished editing up to the discussion

The obesity associated genetic signatures are central to this project, as a means to clarify the relationship between \gls{bmi} and tumour biology.
In this chapter, the previously identified obesity associated genetic signatures from the studies conducted by \citet{Creighton2012} and \citet{Fuentes-Mattei2014} are examined in turn to judge the agreement of these signatures with the sample \gls{bmi}/\gls{bmi} status, as presented in thier results.
After this, novel obesity associated genetic signatures are identified in the Creighton \textit{et al.} (CR) data set and was compared with the signatures from the Creighton \textit{et al.} and the Fuentes-Mattei \textit{et al.} studies.
Lastly, the presence of common genes or pathways associated with obesity in multiple types of cancer is explored using the data sets from \gls{icgc} (\citetalias{ICGC}; \citealp{Zhang2011}).

@@

An obesity metagene was created using the obesity associated genetic signature from the \citet{Creighton2012} study.
There was no description about the normalisation method used by Creighton \textit{et al.} when they first analysed their data, so the popular \gls{rma} method \citep{Irizarry2003} was used to  normalise the CR data set.
The obesity metagene was created in the standardised \gls{rma}-normalised CR data set, and  was plotted above a heatmap with the sample gene expression  to check whether the metagene scores were in accordance with the overall gene expression of the samples (\cref{fig:crmetaheat}).

\begin{figure}[htb]
	\centering
	\includegraphics[page=3,width=0.8\linewidth]{results1/creighton_mg_heatmap1}
	\caption[Obesity metagene from \citet{Creighton2012} study and sample gene expression in CR data] {Heatmap showing the obesity metagene from the \citet{Creighton2012} study with the sample gene expressions of the obesity associated genes in the CR data.
	Level of expression is represented in the top right histogram, where low and high gene expression are colour-coded with blue and red, respectively.
	Each row of the heatmap represents a gene from the obesity associated genetic signature, and each column of the heatmap represents a sample from the CR data.
	The obesity associated metagene scores of the samples are shown as a separate row at top of the heatmap, and the tree diagram of the heirarchical clustering of the genes is shown on the left of the heatmap.
	For clarity, the sign of the metagene scores were reversed in order to match the results from the \citet{Creighton2012} study.}
	\label{fig:crmetaheat}
\end{figure}

As shown in \cref{fig:crmetaheat}, a high obesity associated metagene score  reflects low expression in majority of the genes in the signature, and in contrast, a low obesity associated metagene score reflects high expression in the majority of the genes in the signature.
This was consistent with the reported property of the obesity associated signature by \citet{Creighton2012} (see \cref{ssub:creighton_study}).
To provide further evidence that the obesity associated metagenes were in fact associated with both the \gls{bmi} status and the \gls{bmi} value of the samples, a box plot and a scatter plot were created, respectively (\cref{fig:crmetaboxplot}).

\begin{figure}[htp!]
	\centering
	\includegraphics[page=4,width=0.45\linewidth]{results1/creighton_mg_heatmap1}
	\hfill
	\includegraphics[page=5,width=0.45\linewidth]{results1/creighton_mg_heatmap1}
	\caption[Obesity metagene from the \citet{Creighton2012} study and sample \gls{bmi}/\gls{bmi} status in CR data]{Box plot and scatter plot showing the association of the obesity metagene from the \citet{Creighton2012} study with the sample \gls{bmi} status and \gls{bmi}, respectively, from the CR data.
	In the box plot, the p-values above the groups represent the statistical significance of the association of the obesity metagene with the overweight or obese group compared with the normal weight group.
	The \gls{anova} p-value shows the statistical significance of the association of the obesity metagene with the sample \gls{bmi} groups.
	In the scatter plot, $R^2$- and p-values describe the adjusted coefficient of determination of the regression line and the statistical significance of the linear model used to draw the regression line, respectively.}
	\label{fig:crmetaboxplot}
\end{figure}

\cref{fig:crmetaboxplot} clearly showed that the obesity metagene from the \citet{Creighton2012} study significantly associated with the sample \gls{bmi} status, as well as sample \gls{bmi} value (include p-value and r-squared).
It should be noted here that the obesity associated metagene significantly associated with the samples that were obese, but not with the samples that were overweight.
This was due to the fact that the obesity associated genetic signatures were originally identified from the comparison of the samples that were obese with the samples that were not obese, and therefore the metagene scores were significant with the obese group, but not with the overweight group.
In addition to this, although the regression line in the scatter plot showed statistically significant association with sample \gls{bmi}, the sample \gls{bmi} values seem to be randomly dispersed across the metagene scores.
This suggests that perhaps the obesity metagene from the \citet{Creighton2012} study was not associated with the sample \gls{bmi} as strongly as reported.
\\

\noindent
Now that the association of the obesity metagene from the \citet{Creighton2012} study was established in CR data set, the obesity associated signature was generated in the \gls{icgc} cancer data.
The direction of the obesity associated metagene was checked in the CR data first, so that high metagene scores reflected high sample \gls{bmi} and low metagene scores reflected low sample \gls{bmi} (\cref{fig:crmetaboxplot}).
The transformation matrix was then created in CR data, as described in \cref{sub:svd}.

All of the \gls{icgc} data were normalised as described in \cref{ssub:rna_seq_data}.
Before the transformation matrix was applied to the log$_{10}$-normalised cancer data, the obesity metagene was created from the standardised data or untouched (non-standardised) \gls{icgc} data and compared with one another to determine which was better for the application of the transformation matrix (see \cref{sec:comp_cr_raw_std_icgc}).
From these results, the standardised data was found to be the most suitable to use for matrix transformation.
The transformation matrix was applied to each cancer data set in turn to generate an obesity metagene from each of the data sets.
Each obesity metagene was plotted in a heatmap with the corresponding data set from which the metagene was taken (\cref{fig:crmetaicgc}; \cref{sec:rest_of_the_cr_icgc_cancer_heatmap_results}).
These heatmaps confirmed that the obesity associated metagene derived from the CR data was able to capture the overall gene expression pattern, where the metagene scores reflected the expression levels of the majority of the genes in the signature.
As before, association of the obesity metagene with the sample \gls{bmi} and \gls{bmi} status was checked in their respective cancer data set (\cref{fig:crmetaicgc}; \cref{sec:rest_of_the_cr_icgc_cancer_heatmap_results}).
Out of all the cancer types, only the \gls{blca} data set showed a significant association with the obesity metagene,  and only for the overweight group (not the obese group).
% However, neither the \gls{anova} p-value nor the regression line in the scatter plot were statistically significant, suggesting that this apparent association with the overweight group was not due to the obesity metagene (?).

\begin{figure}[htp!]
	\centering
	\includegraphics[page=3,width=0.8\linewidth]{results1/crtcga_std}\\
	\includegraphics[page=4,width=0.45\linewidth]{results1/crtcga_std}
	\hfill
	\includegraphics[page=5,width=0.45\linewidth]{results1/crtcga_std}
	\caption[Obesity metagene from the \citet{Creighton2012} study in the \acrshort{icgc} \acrshort{blca} data]{Heatmap, box plot and scatter plot showing the association of the obesity metagene from the \citet{Creighton2012} study with the sample gene expression, \gls{bmi} and \gls{bmi} status, respectively, from the \gls{icgc} \gls{blca} data.
	The results for other \gls{icgc} cancer types are shown in \cref{sec:rest_of_the_cr_icgc_cancer_heatmap_results}.
	Scales, p-values and $R^2$-value are as described in previous figures.}
	\label{fig:crmetaicgc}
\end{figure}

There could be several reasons for the apparent lack of association of the metagene with sample \gls{bmi}/\gls{bmi} status.
First, the transformation matrix was derived from the CR microarray data, but the \gls{icgc} cancer data were generated via \gls{rnaseq}.
Though the log$_{10}$-normalisation and standardisation of the data was the most appropriate adjustments to be made to the \gls{rnaseq} data, these adjustments were not equivalent to the \gls{rma} normalisation method that was used on the microarray data.
Secondly, none of the \gls{icgc} cancer originated from breast tumours as in the CR data.
Since the obesity associated signature was identified in the breast cancer data, the signature may be specific to breast cancer and may not be generalisable to other cancer types.
\\

\noindent
To check whether the metagene was specific to breast cancer microarray data, the same transformation matrix was applied to the \gls{nzbc} microarray data from the \citet{Print2016} study (referred to as \gls{nzbc} data hereafter).
\gls{nzbc} data was normalised with the \gls{rma} method and the transformation matrix was applied to the normalised data to obtain the metagene.
The metagene was again compared with the gene expression of the samples with a heatmap and the association of the metagene with the sample \gls{bmi}/\gls{bmi} status was examined with box and scatter plots (\cref{fig:crmetaprint}).

The Creighton \textit{et al.} obesity metagene managed to reflect the overall gene expression of the samples in \gls{nzbc}  data.
However, as with the \gls{icgc} cancer data, the obesity metagene scores did not significantly associate with the sample \gls{bmi} or \gls{bmi} status (\cref{fig:crmetaprint}).
These results confirmed that the lack of association of the obesity metagene from the CR data was not due to the technology in which the data was gathered (mcroarray or \gls{rnaseq}), nor the cancer type in which the transformation matrix was applied to.

\begin{figure}[htp!]
	\centering
	\includegraphics[width=0.8\linewidth,page=3]{results1/cris_cr_trans_meta}\\
	\includegraphics[width=0.45\linewidth,page=4]{results1/cris_cr_trans_meta}
	\hfill
	\includegraphics[width=0.45\linewidth,page=5]{results1/cris_cr_trans_meta}
	\caption[Obesity metagene from \citet{Creighton2012} study in the \gls{nzbc} data]{Heatmap, box plot and scatter plot showing the association of the obesity metagene from the \citet{Creighton2012} study with the sample gene expression, \gls{bmi} and \gls{bmi} status from the \gls{nzbc} data, respectively.
	Scales, p-values and $R^2$-value are as described in previous figures.}
	\label{fig:crmetaprint}
\end{figure}

Taken together, these results suggest that the obesity metagene identified in the study conducted by \citet{Creighton2012} associated with sample \gls{bmi} and \gls{bmi} status only in that particular data set.
The lack of association with \gls{bmi} was not due to the type of technology platform in which the data was gathered, as neither the \gls{icgc} \gls{rnaseq} data nor the \gls{nzbc} microarray data showed a significant association with the obesity metagene.
Furthermore, the obesity metagene was not dependent on the cancer type since the obesity associated genetic signature did not show significant association in either the \gls{icgc} cancer data or in \gls{nzbc} data.

One possible reason why the obesity metagene from the \citet{Creighton2012} study did not show significant association with other data sets could be because the genetic signature was in fact not an obesity specific signature, but a signature that was detected due to another clinical variable (investigated in \cref{sec:creighton_obesity_metagene_new}).
Another reason for this apparent lack of association could be that the genetic signature was too specific to the data and was not a broad obesity associated genetic signature, but an obesity associated signature specific to the samples that were profiled in the Creighton \textit{et al.} publication.

@@

Since the obesity associated genetic signature from the \citet{Creighton2012} study was not associated with sample \gls{bmi} or \gls{bmi} status in other data sets, the obesity associated genetic signature from the \citet{Fuentes-Mattei2014} study (FM) was examined to see whether the FM obesity metagene associated with the sample \gls{bmi} or \gls{bmi} status.
Since the FM data set did not have sample \gls{bmi} information, the FM obesity metagene was not able to be compared with the sample \gls{bmi} or \gls{bmi} status in the original data.
However, the transformation matrix was still applied to other tumour types to see whether the FM obesity metagene associated with sample \gls{bmi}/\gls{bmi} status in those data sets.

Firstly, FM data was normalised with the \gls{rma} method and \gls{svd} was performed on the normalised FM data to obtain the transformation matrix.
The transformation matrix was used to transform the \gls{rma}-normalised CR data to extract the FM obesity metagene scores in the CR data.
The FM obesity metagene scores were compared with gene expression and sample \gls{bmi}/\gls{bmi} status in the CR data, as shown in \cref{fig:fmmetacr}.
Clearly, as with the CR obesity metagene, the FM obesity metagene was reflective of the overall gene expression of the samples, but did not associate with the sample \gls{bmi} or \gls{bmi} status, suggesting that this signature also does not generalise to other data sets.
The transformation matrix was applied to \gls{nzbc} data  as well (\cref{fig:fmmetacris}).
Again, FM obesity metagene scores reflected the gene expression of the samples, did not significantly associate with sample \gls{bmi}/\gls{bmi} status.

Next, the transformation matrix was applied to the \gls{icgc} cancer data and the resulting metagenes were compared with the gene expression and sample \gls{bmi}/\gls{bmi} status.
As evident in \cref{fig:fmmetaicgc}, the FM obesity metagene scores appeared to reflect the overall gene expression of the FM obesity associated genetic signature.
As with all the results in this chapter so far, the FM obesity metagene did not significantly associate with any of the \gls{icgc} cancer data, except in the \gls{blca} data set (\cref{fig:fmmetaicgc}).
The FM obesity metagene significantly associated with the overweight group (but not with the obese group), and also had a significant \gls{anova} p-value.
On the contrary to the association of the metagene with the sample \gls{bmi} status, the FM obesity metagene was not associated with sample \gls{bmi}.
These results suggested that the samples that were overweight in the \gls{blca} data set had similar biological properties as the samples that were obese in the FM data set.
However, due to the fact that the FM metagene lacked association with the sample \gls{bmi} in the \gls{blca} data set and that the metagene did not show any significant association in any other cancer type, it was difficult to determine whether the observed association of the FM obesity metagene with the overweight group of samples was truly reflective of the effect resulted from the FM metagene.
\\

\begin{figure}[htp!]
	\centering
	\includegraphics[page=3,width=0.8\linewidth]{results1/cr_fm_meta}\\
	\includegraphics[page=4,width=0.45\linewidth]{results1/cr_fm_meta}
	\hfill
	\includegraphics[page=5,width=0.45\linewidth]{results1/cr_fm_meta}
	\caption[FM obesity metagene in the CR data]{Heatmap, box plot and scatter plot showing the association of the FM obesity associated metagene with the sample gene expression, \gls{bmi} and \gls{bmi} status from the CR  data, respectively.
	Scales, p-values and $R^2$-value are as described in previous figures.}
	\label{fig:fmmetacr}
\end{figure}

\begin{figure}[htp!]
	\centering
	\includegraphics[page=3,width=0.8\linewidth]{results1/cris_fm_meta}\\
	\includegraphics[page=4,width=0.45\linewidth]{results1/cris_fm_meta}
	\hfill
	\includegraphics[page=5,width=0.45\linewidth]{results1/cris_fm_meta}
	\caption[FM metagene in the \gls{nzbc} data]{Heatmap, box plot and scatter plot showing the association of the FM obesity associated metagene with the sample gene expression, \gls{bmi} and \gls{bmi} status from the \gls{nzbc} data, respectively.
	Scales, p-values and $R^2$-value are as described in previous figures.}
	\label{fig:fmmetacris}
\end{figure}

\begin{figure}[htp!]
	\centering
	\includegraphics[page=3,width=0.8\linewidth]{results1/fm_meta_ICGC}\\
	\includegraphics[page=4,width=0.45\linewidth]{results1/fm_meta_ICGC}
	\hfill
	\includegraphics[page=5,width=0.45\linewidth]{results1/fm_meta_ICGC}
	\caption[FM obesity metagene in the \acrshort{icgc} \acrshort{blca} data]{Heatmap, box plot and scatter plot showing the association of the FM obesity associated metagene with the sample gene expression, \gls{bmi} and \gls{bmi} status from the \acrshort{icgc} \acrshort{blca} data, respectively.
	The results for other \gls{icgc} cancer types are shown in \cref{sec:rest_of_the_fm_icgc_cancer_heatmap_results}.
	Scales, p-values and $R^2$-value are as described in previous figures.}
	\label{fig:fmmetaicgc}
\end{figure}

\noindent
These results showed that FM obesity associated metagene was not generalisable to other cancer data sets, similar to the obesity metagene identified by Creighton \textit{et al.}.
This meant that both the obesity metagenes identified by Creighton \textit{et al.} and Fuentes-Mattei \textit{et al.} may have been too specific to the orignial data set in which the signatures were identified.
Furthermore, there was a possibility that these obesity associated metagenes were not related to obesity, but associated with a different clinical variable that may be closely related to \gls{bmi} (\cref{sec:creighton_obesity_metagene_new}).

@@

\subsection{Identification of obesity associated genetic signatures}
\label{sub:identification_of_obesity_associated_genetic_signatures}

Both the obesity associated metagenes generated from CR data and FM data were able to capture the overall pattern of gene expression of the genetic signature in the samples, but did not associate with sample \gls{bmi} or \gls{bmi} status.
One possible reason for this result could be that the obesity associated genetic signatures from the \citet{Creighton2012} and \citet{Fuente-Mattei2014} studies may not have been truly associated with sample \gls{bmi}  status, but with another clinical variable.
To invevstigate this possibility, obesity associated genetic signatures were identified in CR data after controlling for all the clinical variables in the data set.
FM data set was not used to get obesity associated genetic signatures, as no \gls{bmi} information was available for the samples in FM data.

Firstly, to determine whether the identified obesity associated genetic signature was similar to the signature that Creighton \textit{et al.} had originally found, the identified \glspl{deg} were compared with the original genetic signature identified by Creighton \textit{et al.}.
\citet{Creighton2012} originally found their obesity associated genetic signature by doing a gene expression analysis between the breast tumour samples from the obese and non-obese patients, and from this list, they selected the 799 significant \glspl{deg} with log$_2$-fold change greater than 1.2.

>>>>

The obesity associated genetic signatures are central to this project, as a means to clarify the relationship between \gls{bmi} and tumour biology.
In this chapter, the previously identified obesity associated genetic signatures from the studies conducted by \citet{Creighton2012} and \citet{Fuentes-Mattei2014} are examined in turn to judge the agreement of these signatures with the patient \gls{bmi} and \gls{bmi} status, as presented in their results.
After this, novel obesity associated genetic signatures are identified in the Creighton \textit{et al.} (CR) data set and was compared with the obesity associated genetic signatures from the \citet{Creighton2012} and the \citet{Fuentes-Mattei2014}  studies.
Lastly, the presence of common genes or pathways associated with obesity in multiple types of cancer is explored using the data sets from \gls{icgc} (\citetalias{ICGC}; \citealp{Zhang2011}).

@@

An obesity metagene was created using the obesity associated genetic signature from the \citet{Creighton2012} study.
There was no description about the normalisation method used by Creighton \textit{et al.} when they first analysed their data, so the more popular \gls{rma} method  was used to  normalise the CR data set \citep{Irizarry2003}.
The obesity metagene was created in the standardised \gls{rma} normalised CR data set, and  was plotted above a heatmap with the sample gene expression  to check whether the metagene scores were in accordance with the overall gene expression of the samples (\cref{fig:crmetaheat}).

\begin{figure}[tb!]
	\centering
	\includegraphics[page=3,width=0.8\linewidth]{results1/creighton_mg_heatmap1}
	\caption[Obesity metagene from \citet{Creighton2012} study and sample gene expression in CR data] {Heatmap showing the obesity metagene from the \citet{Creighton2012} study with the sample gene expressions of the obesity associated genes in the CR data.
	Level of expression is represented in the top right histogram, where low and high gene expression are colour-coded with blue and red, respectively.
	Each row of the heatmap represents a gene from the obesity associated genetic signature, and each column of the heatmap represents a sample from the CR data.
	The obesity associated metagene scores of the samples are shown as a separate row at top of the heatmap, and the tree diagram of the heirarchical clustering of the genes is shown on the left of the heatmap.
	For clarity, the sign of the metagene scores were reversed in order to match the results from the \citet{Creighton2012} study.}
	\label{fig:crmetaheat}
\end{figure}

As shown in \cref{fig:crmetaheat}, a high obesity associated metagene score  reflects low expression in majority of the genes in the signature, and in contrast, a low obesity associated metagene score reflects high expression in the majority of the genes in the signature.
This was consistent with the reported property of the obesity associated genetic signature by \citet{Creighton2012} (see \cref{ssub:creighton_study}).
To provide further evidence that the obesity metagenes were in fact associated with both the \gls{bmi} status and the \gls{bmi} value of the patients, a box plot and a scatter plot were created, respectively (\cref{fig:crmetaboxplot}).

\begin{figure}[tb!]
	\centering
	\includegraphics[page=4,width=0.45\linewidth]{results1/creighton_mg_heatmap1}
	\hfill
	\includegraphics[page=5,width=0.45\linewidth]{results1/creighton_mg_heatmap1}
	\caption[Obesity metagene from the \citet{Creighton2012} study and sample \gls{bmi}/\gls{bmi} status in CR data]{Box plot and scatter plot showing the association of the obesity metagene from the \citet{Creighton2012} study with the sample \gls{bmi} status and \gls{bmi}, respectively, from the CR data.
	In the box plot, the p-values above the groups represent the statistical significance of the association of the obesity metagene with the overweight or obese group compared with the normal weight group.
	The \gls{anova} p-value shows the statistical significance of the association of the obesity metagene with the sample \gls{bmi} groups.
	In the scatter plot, $R^2$- and p-values describe the adjusted coefficient of determination of the regression line and the statistical significance of the linear model used to draw the regression line, respectively.}
	\label{fig:crmetaboxplot}
\end{figure}

\cref{fig:crmetaboxplot} clearly showed that the obesity metagene from the \citet{Creighton2012} study significantly associated with the obese group of patients, as well as patient \gls{bmi} value.
It should be noted here that the obesity metagene significantly associated with the samples from the patients that were obese, but not with the sample from the patients that were overweight.
This was due to the fact that the obesity associated genetic signature was originally identified from the comparison of the patients that were obese with the patients that were not obese, and therefore the obesity metagene scores were significant with the obese group, but not with the overweight group.
Another thing to note here was that even though  the regression line in the scatter plot showed statistically significant association with the patient \gls{bmi}, patient \gls{bmi} values seem to be randomly dispersed across the obesity metagene scores.
This suggests that perhaps the obesity metagene from the \citet{Creighton2012} study may not be associated with the patient \gls{bmi} as strongly as reported.
\\

\noindent
Now that the association of the obesity metagene from the \citet{Creighton2012} study was established in the CR data set, the Creighton \textit{et al.} obesity metagene was generated in the \gls{icgc} cancer data.
The direction of the obesity metagene was checked in the CR data first, so that high metagene scores reflected high patient \gls{bmi} and low metagene scores reflected low patient \gls{bmi} (\cref{fig:crmetaboxplot}).
The transformation matrix was then created in the \gls{rma} normalised CR data, as described in \cref{sub:svd}.

All of the \gls{icgc} \gls{rnaseq} data were normalised as described in \cref{ssub:rna_seq_data}.
Before the transformation matrix was applied to the log$_{10}$-normalised cancer data, the obesity metagene was created from the standardised data or untouched (non-standardised) \gls{icgc} data and compared with one another to determine which data format was better for the application of the transformation matrix (\cref{sec:comp_cr_raw_std_icgc}).
From these results, the standardised data was found to be the most suitable format for the application  of the transformation matrix.
The transformation matrix was applied to each cancer data set in turn to generate an obesity metagene from each of the data sets.
Each obesity metagene was plotted in a heatmap with the corresponding \gls{icgc} data set from which the obesity metagene was generated from (\cref{fig:crmetaicgc}; \cref{sec:rest_of_the_cr_icgc_cancer_heatmap_results}).
These heatmaps confirmed that the obesity metagene was able to capture the overall gene expression pattern in all of the \gls{icgc} cancer data, where the metagene scores reflected the expression levels of the majority of the genes in the signature.
As before, association of the obesity metagene with the patient \gls{bmi} and \gls{bmi} status was examined in their respective cancer data set (\cref{fig:crmetaicgc}; \cref{sec:rest_of_the_cr_icgc_cancer_heatmap_results}).
Out of all the cancer types, only the \gls{blca} data set showed a significant association with the obesity metagene,  and only for the overweight group (not the obese group).

\begin{figure}[htp!]
	\centering
	\includegraphics[page=3,width=0.8\linewidth]{results1/crtcga_std}\\
	\includegraphics[page=4,width=0.45\linewidth]{results1/crtcga_std}
	\hfill
	\includegraphics[page=5,width=0.45\linewidth]{results1/crtcga_std}
	\caption[Obesity metagene from the \citet{Creighton2012} study in the \acrshort{icgc} \acrshort{blca} data]{Heatmap, scatter plot and box plot showing the association of the obesity metagene from the \citet{Creighton2012} study with the sample gene expression, patient \gls{bmi} and \gls{bmi} status, respectively, from the \gls{icgc} \gls{blca} data.
	The results for the other \gls{icgc} cancer types are shown in \cref{sec:rest_of_the_cr_icgc_cancer_heatmap_results}.
	Scales, p-values and $R^2$-value are as described in previous figures.}
	\label{fig:crmetaicgc}
\end{figure}

There could be several reasons for the apparent lack of association of the obesity metagene with patient \gls{bmi} in most of the \gls{icgc} data.
First, the transformation matrix was derived from the CR microarray data, but the \gls{icgc} cancer data were generated via \gls{rnaseq}.
Though the log$_{10}$-normalisation and standardisation of the data was the most appropriate adjustments to be made to the \gls{rnaseq} data, these adjustments were not equivalent to the \gls{rma} normalisation method that was used on the microarray data.
Secondly, none of the \gls{icgc} cancer data in this project originated from breast tumours as in the CR data.
Since the obesity associated genetic signature was identified in the breast cancer data, the signature may be specific to breast cancer and may not be generalisable to other cancer types.
\\

\noindent
To check whether the signature was specific to breast cancer microarray data, the same transformation matrix used in the \gls{icgc} data was applied to the breast cancer microarray data from the \citet{Print2016} study (referred to as \gls{nzbc} data hereafter).
\gls{nzbc} data was normalised with the \gls{rma} method and the transformation matrix was applied to the normalised data to obtain the Creighton \textit{et al.} obesity metagene in the \gls{nzbc} data set.
The generated obesity metagene was again compared with the gene expression of the samples with a heatmap and the association of the metagene with the patient \gls{bmi} and \gls{bmi} status was examined with box and scatter plots, respectively  (\cref{fig:crmetaprint}).

The Creighton \textit{et al.} obesity metagene managed to reflect the overall gene expression of the samples in the \gls{nzbc}  data (\cref{fig:crmetaprint}).
However, as with the \gls{icgc} cancer data, the obesity metagene scores did not significantly associate with patient \gls{bmi} or \gls{bmi} status (\cref{fig:crmetaprint}).
These results confirmed that the lack of association of the obesity associated genetic signature from the Creighton \textit{et al.} study was not due to the technology in which the data was gathered (microarray or \gls{rnaseq}), nor the cancer type in which the genetic signature was derived from.

\begin{figure}[htp!]
	\centering
	\includegraphics[width=0.8\linewidth,page=3]{results1/cris_cr_trans_meta}\\
	\includegraphics[width=0.45\linewidth,page=4]{results1/cris_cr_trans_meta}
	\hfill
	\includegraphics[width=0.45\linewidth,page=5]{results1/cris_cr_trans_meta}
	\caption[Obesity metagene from the \citet{Creighton2012} study in the \gls{nzbc} data]{Heatmap, scatter plot and box plot showing the association of the obesity metagene from the \citet{Creighton2012} study with the sample gene expression, patient \gls{bmi} and \gls{bmi} status, respectively, from the \gls{nzbc} data.
	Scales, p-values and $R^2$-value are as described in previous figures.}
	\label{fig:crmetaprint}
\end{figure}

Taken together, these results suggest that the obesity associated genetic signature identified in the study conducted by \citet{Creighton2012} correlated with patient \gls{bmi} and \gls{bmi} status only in the CR data set.
The lack of association with patient \gls{bmi} in other cancer data sets was not due to the type of technology platform in which the data was gathered, as neither the \gls{icgc} \gls{rnaseq} data nor the \gls{nzbc} microarray data showed significant association of the obesity metagene with the patient \gls{bmi}.
Furthermore, the obesity associated genetic signature was not dependent on the cancer type in which it was generated from, since the obesity metagene did not show significant association in either the \gls{icgc} cancer data or in the \gls{nzbc} data.

One possible reason why the obesity metagene from the \citet{Creighton2012} study did not show significant association with other data sets could be because the genetic signature was in fact not an obesity specific signature, but a signature that was detected due to another clinical variable (investigated in \cref{sec:creighton_obesity_metagene_new}).
Another reason for this apparent lack of association could be that the obesity associated genetic signature was too specific to the CR data and was not a broad obesity associated genetic signature, but an obesity associated signature that was specific to the patient cohort that was profiled in the \citet{Creighton2012} publication.
% Another reason for this apparent lack of association could be that the genetic signature was too specific to the data and was not a broad obesity associated genetic signature, but an obesity associated signature that was specific to the samples that were profiled in the Creighton \textit{et al.} publication.

@@

The obesity associated genetic signature from the \citet{Creighton2012} study was not associated with patient \gls{bmi} or \gls{bmi} status in majority of the cancer data sets, so the obesity associated genetic signature from the \citet{Fuentes-Mattei2014} study (FM) was examined to see whether this obesity metagene was able to significantly associate with the patient \gls{bmi} and \gls{bmi} status across different  data sets.
Since the FM data set did not have patient \gls{bmi} information, the FM obesity metagene was not able to be compared with the patient \gls{bmi} or \gls{bmi} status in the original FM data.
Nevertheless, the transformation matrix was still generated in the FM data and applied first to the microarray data (CR and \gls{nzbc} data sets) then to the \gls{icgc} data sets to see whether the FM obesity metagene associated with patient \gls{bmi} and \gls{bmi} status in these data sets.
% However, the transformation matrix was generated in the FM data and applied to other cancer data sets  to see whether the FM obesity metagene associated with patient \gls{bmi} and \gls{bmi} status in those data sets.

The FM data was normalised with the \gls{rma} method and \gls{svd} was performed on the normalised FM data to obtain the transformation matrix.
The transformation matrix was used to transform the \gls{rma} normalised CR data to extract the FM obesity metagene scores in the CR data.
The FM obesity metagene scores were compared with sample gene expression levels, patient \gls{bmi} and \gls{bmi} status in the CR data (\cref{fig:fmmetacr}).
Clearly, as with the CR obesity metagene, the FM obesity metagene was reflective of the overall gene expression of the samples, but did not associate with patient \gls{bmi} or \gls{bmi} status, suggesting that this signature also does not generalise to other data sets.
The transformation matrix was then applied to the \gls{nzbc} microarray  data.
Again, FM obesity metagene scores reflected the gene expression of the samples, but did not significantly associate with patient \gls{bmi} or \gls{bmi} status (\cref{fig:fmmetacris}).

\begin{figure}[htp!]
	\centering
	\includegraphics[page=3,width=0.8\linewidth]{results1/cr_fm_meta}\\
	\includegraphics[page=4,width=0.45\linewidth]{results1/cr_fm_meta}
	\hfill
	\includegraphics[page=5,width=0.45\linewidth]{results1/cr_fm_meta}
	\caption[FM obesity metagene in the CR data]{Heatmap, scatter plot and box plot showing the association of the FM obesity metagene with the sample gene expression, patient \gls{bmi} and \gls{bmi} status, respectively, from the CR  data.
	Scales, p-values and $R^2$-value are as described in previous figures.}
	\label{fig:fmmetacr}
\end{figure}

\begin{figure}[htp!]
	\centering
	\includegraphics[page=3,width=0.8\linewidth]{results1/cris_fm_meta}\\
	\includegraphics[page=4,width=0.45\linewidth]{results1/cris_fm_meta}
	\hfill
	\includegraphics[page=5,width=0.45\linewidth]{results1/cris_fm_meta}
	\caption[FM metagene in the \gls{nzbc} data]{Heatmap, scatter plot and box plot showing the association of the FM obesity associated metagene with the sample gene expression, patient  \gls{bmi} and \gls{bmi} status, respectively, from the \gls{nzbc} data.
	Scales, p-values and $R^2$-value are as described in previous figures.}
	\label{fig:fmmetacris}
\end{figure}

Next, the transformation matrix was applied to the \gls{icgc} cancer data and the resulting FM metagene was compared with the sample gene expression, patient \gls{bmi} and \gls{bmi} status in each of the \gls{icgc} cancer data.
As evident in \cref{fig:fmmetaicgc,sec:rest_of_the_fm_icgc_cancer_heatmap_results}, the FM obesity metagene scores appeared to reflect the overall gene expression of the FM obesity associated genetic signature.
As with all the results in this chapter so far, the FM obesity metagene did not significantly associate with any of the \gls{icgc} cancer data, except in the \gls{blca} data set (\cref{fig:fmmetaicgc}; \cref{sec:rest_of_the_fm_icgc_cancer_heatmap_results}).
The FM obesity metagene significantly associated with the overweight group (but not with the obese group), and also had a significant \gls{anova} p-value \cref{fig:fmmetaicgc}.
On the contrary to the association of the metagene with the patient \gls{bmi} status, the FM obesity metagene was not associated with the patient \gls{bmi}.
These results suggested that the samples from the patients that were overweight in the \gls{blca} data set had similar biological properties as the samples taken from the patients that were obese in the FM data set.
However, due to the fact that the FM obesity metagene lacked association with the patient \gls{bmi} in the \gls{blca} data set and that the metagene did not show any significant association in any of the other \gls{icgc} cancer types, it was difficult to determine whether the observed association of the FM obesity metagene with the overweight group was truly reflective of the quality of the FM metagene or observed by chance.

\begin{figure}[htp!]
	\centering
	\includegraphics[page=3,width=0.8\linewidth]{results1/fm_meta_ICGC}\\
	\includegraphics[page=4,width=0.45\linewidth]{results1/fm_meta_ICGC}
	\hfill
	\includegraphics[page=5,width=0.45\linewidth]{results1/fm_meta_ICGC}
	\caption[FM obesity metagene in the \acrshort{icgc} \acrshort{blca} data]{Heatmap, scatter plot and box plot showing the association of the FM obesity metagene with the sample gene expression, patient \gls{bmi} and \gls{bmi} status, respectively, from the \acrshort{icgc} \acrshort{blca} data.
	The results for other \gls{icgc} cancer types are shown in \cref{sec:rest_of_the_fm_icgc_cancer_heatmap_results}.
	Scales, p-values and $R^2$-value are as described in previous figures.}
	\label{fig:fmmetaicgc}
\end{figure}

These results showed that the obesity associated genetic signature from the \citet{Fuentes-Mattei2014} study was not generalisable to other cancer data sets, similar to the obesity signature identified by Creighton \textit{et al.}.
This meant that both the Creighton \textit{et al.} and Fuentes-Mattei \textit{et al.} obesity metagenes may have been too specific to the orignial data set in which the signatures were identified in.
Furthermore, there was a possibility that these obesity associated metagenes were not related to obesity, but associated with a different clinical variable that may be closely related to \gls{bmi} (\cref{sec:creighton_obesity_metagene_new}).

@@

\subsection{Identification of novel obesity associated genetic signatures}
\label{sub:identification_of_obesity_associated_genetic_signatures}

Both the obesity associated genetic signatures identfied from the \citet{Creighton2012} and \citet{Fuentes-Mattei2014} studies were able to capture the overall gene expression patterns  of the genetic signatures of the samples, but did not associate with patient \gls{bmi} or \gls{bmi} status in majority of the cancer data sets.
One possible reason for this result could be that the obesity associated genetic signatures from the \citet{Creighton2012} and \citet{Fuentes-Mattei2014} studies may not have been truly associated with patient \gls{bmi}, but with another clinical variable.
To invevstigate this possibility, novel obesity associated genetic signatures were identified in the CR data after controlling for all the clinical variables in the data set.
The FM data set was not used to get the obesity associated genetic signatures, as no \gls{bmi} information was available for the patients in the FM data.
